The effect of climate forcing on numerical simulations of the Cordilleran Ice Sheet at the last glacial maximum

### 2013-06-13 sections/intro.tex
Polish the introduction

In numerical models of the former North American ice sheets, various climate forcing methods have been used, ranging from simplistic parametrizations to coupling to General Circulation Models (GCM). Arguably the most physical way to force an ice sheet model in the past is to couple it with a GCM \citep{yoshimori-etal-2001,calov-etal-2002,abeouchi-etal-2007,charbit-etal-2013}. However due to the high computational cost of this exercise only GCMs of coarser resolution and reduced complexity can achieve it.

More commonly GCM output from palæo-climate simulations is used as a forcing \citep{huybrechts-tsiobbel-1996}. To produce time-dependant forcing, GCM is often interpolated in time between different palæo-climate simulations, either linearly \citep{charbit-etal-2002} or more commonly using a wieghting function based on ice-core records \citep{marshall-clarke-1999,tarasov-peltier-2004,zweck-huybrechts-2005,gregoire-etal-2012}. GCM palæo-climate simulations typically rely on ice-sheet reconstructions of the type of the ICE-4G \citep{peltier-1994} for their surface topography boundary condition. In turn, ice-sheet models based on these simulations generally reproduce LGM ice extents closely consistent with the reconstruction.

Other methods include parametrization of climate forcing based on elevation \citep{robert-1991}, geographical coordinates \citep{fastook-prentice-1994,johnson-fastook-2002} and temperature offsets applied to a present-day reference climate \citep{bintanja-etal-2005}.

@@

Our area of focus for the present study is the Cordilleran Ice Sheet in north-western North America. In a numerical modelling perspective, it is one of the least studied paleo-ice sheets on Earth, where nethertheless significant geomorphological data is available \citep{jackson-clague-1991}. Although generally not the primary target of these simulations, the Cordilleran Ice Sheet has been previously modelled as part of the North American ice sheet complex \citep{marshall-clarke-1999,calov-etal-2002,tarasov-peltier-2004,bintanja-etal-2005,gregoire-etal-2012} and within a larger-scale Arctic or Global numerical domain \citep{huybrechts-tsiobbel-1996,charbit-etal-2002,johnson-fastook-2002,zweck-huybrechts-2005,abeouchi-etal-2007,charbit-etal-2013}.

If these studies reproduce well the magnitude of the North American glaciation at Last Glacial Maximum, there is a general tendency for studies that do not undirectly rely on reconstructions of the type of ICE-4G \citep{johnson-fastook-2002,bintanja-etal-2005,huybrechts-tsiobbel-1996,charbit-etal-2002,marshall-clarke-1999,tarasov-peltier-2004,zweck-huybrechts-2005,gregoire-etal-2012} to predict excessive ice cover in northern Yukon Territory and Alaska, regions that have not been covered by ice sheets throughout the Pleistocene \citep{dukrodkin-1999,kaufman-manley-2004}.

Here we use a Parallel Ice Sheet Model (PISM) \citep{web:pism} to simulate the entire Cordilleran Ice Sheet at the Last Glacial Maximum (LGM). We force our model with multiple climate datasets and compare our results to the mapped LGM ice sheet margin from \citet{kleman-etal-2010}. See \citet{quiquet-etal-2012} for a similar comparison on Greenland. To avoid circular dependance on prescribed ice sheet geometries, and allow high spatial resolution in our climate forcing, we use present-day climate data and given temperature offsets in a manner similar to \citet{bintanja-etal-2005}. To our knowledge, this is the first modelling study which specifically focus on the Cordilleran Ice Sheet since the one by \citep{robert-1991}, who focused specifically on its southern margin.

@@

>>>>

Numerical models of the former North American ice sheets have used various climate forcing methods, ranging from simplistic parametrizations to coupling to General Circulation Models (GCM). The latter is arguably the most physically sound way to force an ice sheet model in the past and has proven effective \citep{yoshimori-etal-2001,calov-etal-2002,abeouchi-etal-2007,charbit-etal-2013}. However the computational demand of GCMs is so that only models of reduced complexity and coarser resolution can run on time-scales of ice sheet growth and decay.

Another method consist in using atmospheric fields produced in uncoupled palæo-climate simulations \citep{huybrechts-tsiobbel-1996}. To produce a time-dependant forcing, the GCM output is interpolated in time between different simulations \citep{charbit-etal-2002} or weighted based on ice-core records \citep{marshall-clarke-1999,tarasov-peltier-2004,zweck-huybrechts-2005,gregoire-etal-2012}. These palæo-climate simulations use ice sheet reconstructions such as the ICE-4G \citep{peltier-1994} as topographic boundary condition, which results in subsequently modelled ice cover consistent with these reconstructions.
\ask[noline]{I am thinking whether I should extend this literature list to papers not concerning North America?}

@@

Avoiding this circular dependence unfortunately goes with simplistic assumptions on Earth's past climate. \citet{robert-1991} use a parametrization of glacier surface mass balance based on elevation. \citet{fastook-prentice-1994} use a parametrization of climate forcing based on elevation and distance from a climatic pole initially prepared for Antarctica \citep{johnson-fastook-2002}. Finally, \citet{bintanja-etal-2005} apply time-dependant temperature offsets to a reference present-day climate obtained from climate reanalysis data. We use a similar method.

@@

Our area of focus for the present study is the Cordilleran Ice Sheet in north-western North America (Figure 1). In a numerical modelling perspective, it is one of the least studied palæo-ice sheets of the Northern Hemisphere, where nevertheless significant geomorphological data is available \citep{jackson-clague-1991}. Although generally not the primary target of these simulations, the Cordilleran Ice Sheet has been previously modelled as part of the North American ice sheet complex \citep{marshall-clarke-1999,calov-etal-2002,tarasov-peltier-2004,bintanja-etal-2005,gregoire-etal-2012} and within a larger-scale Arctic or Global numerical domain \citep{huybrechts-tsiobbel-1996,charbit-etal-2002,johnson-fastook-2002,zweck-huybrechts-2005,abeouchi-etal-2007,charbit-etal-2013}.

While these studies reproduce well the magnitude of the North American glaciation at Last Glacial Maximum, they tend to predict excessive ice cover in parts of northern Yukon Territory and Alaska that have remained ice-free throughout the Pleistocene \citep{dukrodkin-1999,kaufman-manley-2004}.

Here we use a Parallel Ice Sheet Model \citep[PISM,][]{web:pism} to simulate the entire Cordilleran Ice Sheet at the Last Glacial Maximum (LGM). We force our model with multiple climate datasets and compare our results to the mapped LGM ice sheet margin from \citet{kleman-etal-2010}. See \citet{quiquet-etal-2012} for a similar comparison on Greenland. To avoid any dependence on ice-sheet reconstructions and allow high spatial resolution in our climate forcing, we use present-day climate data and given temperature offsets in a manner similar to \citet{bintanja-etal-2005}. To our knowledge, this is the first modelling study which specifically focus on the Cordilleran Ice Sheet since the one by \citep{robert-1991}, who focused specifically on its southern margin.
